A generalized quadratic estimate for random field nonstationarity \thanksref{code}

### 2016-02-28 paper/paper.tex
added a bit more detail to the proof of the eulerian characterization of the change of varibles formula

In this paper, we attempt to shed light on a new class of nonstationary random fields which exhibit, what we call, {\em local invariant nonstationarity}. Formally defined in Section \ref{section: local invariant}, a locally invariant nonstationary random field $\{Z(\bs x)\colon \bs x\in \Bbb R^d\}$ has the feature that the covariance function can be written in the following form,
\begin{align*}
\text{cov}(Z(\bs x), Z(\bs y))=K(\bs x-\bs y, \bs \theta(\bs x)-\bs \theta(\bs y)),
\end{align*}
where $K(\cdot, \cdot)\colon \Bbb R^d\times \Bbb R^d \rightarrow \Bbb R$ and $\bs\theta(\cdot) \colon \Bbb R^d\rightarrow \Bbb R^d$ is a vector field which characterizes the nonstationarity in $Z(\bs x)$. We argue that the locally invariant property has a special interaction with a new  generalized quadratic estimate (derived in Section \ref{Section: the quad est}) which extends an estimate originally developed for gravitational lensing of the Cosmic Microwave Background in Cosmology \cite{hu2001mapping, hu2002mass}. The nature of this interaction not only encourages low estimation bias but also enables accurate (and fast) Frequentist mean square error quantification of the nonstationarity characterized by $\bs \theta(\bs x)$. These quadratic estimates are interesting in their own right, as they detect and estimate nonstationarity by probing correlation among Fourier frequencies, the absence of which is the characterizing feature of weak stationarity (by Bochner's Theorem). Moreover, this generalized quadratic estimate can be computed with a Fourier characterization that runs in $\mathcal O(n\log n)$ time and $O(n)$ storage when observing the field on a uniform grid of size $n$ in $\Bbb R^d $. Finally, the work presented here partially addresses two unanswered questions that arise in two important bodies of work on nonstationarity: 1) estimating the spatially varying phase in the models analyzed by Dahlhaus \cite{dahlhaus1997fitting,dahlhaus2000likelihood} and 2) how to extend the quadratic estimate of dark matter from the Cosmic Microwave Background to more general nonstationary random fields.

The seminal work of Dahlhaus in the time series literature \cite{dahlhaus1997fitting,dahlhaus2000likelihood}  is  an example of a collection of results that hint at a more general statistical theory of nonstationarity.  Dahlhaus develops asymptotic theory for a particular class of nonstationary random fields modeled by a spatially (or temporally) varying spectral density. These random fields were originally developed for time series \cite{priestley1965evolutionary,priestley1981spectral} and have the form 
\begin{align}
\label{eq: intro nonstat phase model}
Z(\bs x)= \int_{\Bbb R^d} e^{i\bs x\cdot \bs k} A(\bs k, \bs x)\sqrt{C_{\bs k}}\frac{dW_{\bs k}}{(2\pi)^{d/2}}
\end{align} 
where  $C_{\bs k}$ is a spectral density, $dW_{\bs k}$ is an orthogonal increment random complex measure that satisfies $E|dW_{\bs k}|^2 =  d{\bs k}$ (see \cite{Gikhman_Skorokhod_book_v1} for details on random measures) and $A({\bs k},{\bs x})$ represents a spatial (or temporal) modulation of $\sqrt{C_{\bs k}}$. 
Dahlhaus proved consistency results for estimating the squared modulus $|A({\bs k},{\bs x})|^2$ when estimation is done  by maximizing a weighted sum of local likelihood functions. Left unanswered, however, is the question of estimating the phase of $A({\bs k},{\bs x})$. In Section \ref{section: NPhase}, we partly resolve this question by showing that a generalized quadratic estimate can be used to estimate a pure phase modulation $A({\bs k},{\bs x})=\exp({iB(\bs k,\bs x)})$ where $B(\bs k,\bs x)$ is a function taking values in $\Bbb R$ and is a separable function of $\bs x$ and $\bs k$. Indeed, this nonstationary phase model has the locally invariant property defined in this paper and, as such, can be accurately estimated (under certain conditions on $B(\bs k,\bs x)$) with the generalized quadratic estimate derived in this paper.

Another important development in the statistical theory of nonstationarity comes from recent gravitational lensing studies of the Cosmic Microwave Background (CMB) \citep{das2011detection,van2012measurement, planck2013lensing, Polarbear2014, planck2015lensing}. In Cosmology, gravitational lensing describes the distortion of photon trajectories due to density fluctuations of intervening dark matter. These density fluctuations affect the CMB observations by introducing small nonstationarities in the original isotropic random field model of the CMB. The state-of-the-art estimator of lensing, the quadratic estimator developed by Hu and Okamoto \cite{hu2001mapping, hu2002mass}, has become an incredibly successful tool for probing the nature of dark matter, understanding cosmic structure and constraining cosmological parameters. What is so surprising about this estimate is that is has small bias. This is due to a delicate cancellation of terms in an infinite Taylor expansion of the lensing effect. Unfortunately there has been no clear explanation as to why this cancellation occurs and whether or not it exists in other models of nonstationarity. In this paper, we argue that this cancellation is due to the fact that the lensing-induced nonstationarity is locally invariant. Moreover, the generalized quadratic estimator developed here extends the lensing estimator to more general settings. Indeed, much of techniques we use to derive the generalized quadratic estimator are inspired by---and closely follow---those used by Hu and Okamoto \cite{hu2001mapping, hu2002mass}.  The point of this paper, in contrast to the work of Hu and Okamoto, is to identify the cause of the delicate Taylor series cancellation and extend the benefits of the lensing estimator to a larger class of nonstationary random fields available to general practitioners of spatial statistics.

@@

% --------------------------------------------------------
%  claim
%---------------------------------------------------------
\begin{claim}[\textbf{Any pair of spectral densities with the same integral are attainable}]
\label{claim any two spec are attainable}
Let $d\geq 1$ be an integer, $t_0 > 0$ be a real number and $C_{\bs k}$, $\tilde C_{\bs k}$ be two spectral densities on $\Bbb R^d$ with finite second moments such that $\sigma^2 = \int_{\Bbb R^d}  C_{\bs k}d\bs k = \int_{\Bbb R^d} \tilde C_{\bs k}d\bs k$. Then there exists a vector field $\bs \eta_{\bs k}\colon \Bbb R^d \rightarrow \Bbb R^d$ which is $L_2(\Bbb R^d)$ integrable with respect to $C_{\bs k}d\bs k$ and  generates a one dimensional curve of spectral measures  $\{\lambda^{(t)}: t\in [0,t_0]\}$, defined by 
\begin{align}
\lambda^{(t)} := (\bs k + t\bs \eta_{\bs k})\,\sharp\, C_{\bs k}
\end{align}
with endpoints $\lambda^{(0)}=C_{\bs k}d\bs k$ and $\lambda^{(t_0)}=\tilde C_{\bs k}d\bs k$,
such that  $\{\lambda^{(t)}:t\in[0,t_0] \}\subset \mathscr C^{C,\bs\eta}$. In particular there exists a nonstationary phase random field model for which  $C_{\bs k}$ and $\tilde C_{\bs k}$ are both locally attainable.
Moreover,  $\{\lambda^{(t)}: t\in [0,t_0]\}$ is a $L_2$-Wasserstein geodesic path within the class of absolutely continuous spectral measures (with total mass $\sigma^2$ and finite second moments) and each measure $\lambda^{(t)}$ has a density $C^{(t)}_{\bs k}$ with respect to Lebesgue measure on $\Bbb R^d$ which (weakly) satisfies 
\begin{align}
\label{div flow for geo}
\partial_t C^{(t)}_{\bs k} +\text{\rm div}(\bs \eta^{(t)}_{\bs k}C^{(t)}_{\bs k}) = 0
\end{align}
for all $t\in [0,t_0]$ where $\bs \eta^{(t)}_{\bs k}:=\bs \eta_{T_t^{-1}(\bs k)}$ and $T_t(\bs k):= \bs k + t\bs \eta_{\bs k}$.
\end{claim}
% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By standard optimal transport theory (see \cite{villani2003topics} for example) the assumptions on $C_{\bs k}$ and $\tilde C_{\bs k}$ guarantee the existence of a convex function $\psi_{\bs k}:\Bbb R^d \rightarrow \Bbb R$ such that $\nabla \psi_{\bs k}$ is the optimal transport from $C_{\bs k}d\bs k$ to $\tilde C_{\bs k}d\bs k$.  Let $\bs \eta_{\bs k}:= \frac{1}{t_0}(\nabla \psi_{\bs k}-\bs k)$ so that
\begin{align}
\label{geo eq number 1}
\lambda^{(t)}:=(\bs k + t\bs\eta_{\bs k})\,\sharp\, C_{\bs k} = \big((1-\textstyle\frac{t}{t_0})\bs k + \textstyle\frac{t}{t_0}\nabla\psi_{\bs k}\big)\,\sharp\, C_{\bs k}. 
\end{align}
The particular form of the right hand side of (\ref{geo eq number 1})  implies each measure $\lambda^{(t)}$ has a density $C_{\bs k}^{(t)}$ with respect to Lebesgue measure and the path of measures $\{\lambda^{(t)}\colon t\in[0,t_0] \}$ forms an $L_2$-Wasserstein geodesic with endpoints $C_{\bs k}d\bs k$ and $\tilde C_{\bs k}d\bs k$ at $t=0$ and $t=t_0$ respectively (see Proposition 5.9 in \cite{villani2003topics}). Moreover, by Theorem 5.34 of \cite{villani2003topics},  $C^{(t)}_{\bs k}$ weakly satisfies (\ref{div flow for geo}).
By  setting $\bs A=t\bs I_d$ in (\ref{eq: local spectral density for nonstationary phase}) on has
\[
\int  \exp\!\big(i(\bs x-\bs y)\!\cdot\! (\bs k + t\bs\eta_{\bs k})\big)\, C_{\bs k}\frac{d\bs k}{(2\pi)^d} 
=\int \exp\!\big(i(\bs x-\bs y)\!\cdot\! \bs\omega\big)\,  C^{(t)}_{\bs \omega}\frac{d\bs \omega}{(2\pi)^d}
\]
which implies that for each $t\in[0,t_0]$ the measure $\lambda^{(t)}$ is a locally attainable spectral measure.
\end{proof}

>>>>

In this paper, we attempt to shed light on a new class of nonstationary random fields which exhibit, what we call, {\em local invariant nonstationarity}. Formally defined in Section \ref{section: local invariant}, a locally invariant nonstationary random field $\{Z(\bs x)\colon \bs x\in \Bbb R^d\}$ has the feature that the covariance function can be written in the following form,
\begin{align*}
\text{cov}(Z(\bs x), Z(\bs y))=K(\bs x-\bs y, \bs \theta(\bs x)-\bs \theta(\bs y)),
\end{align*}
where $K(\cdot, \cdot)\colon \Bbb R^d\times \Bbb R^d \rightarrow \Bbb R$ and $\bs\theta(\cdot) \colon \Bbb R^d\rightarrow \Bbb R^d$ is a vector field which characterizes the nonstationarity in $Z(\bs x)$. We argue that the locally invariant property has a special interaction with a new  generalized quadratic estimate (derived in Section \ref{Section: the quad est}) which extends an estimate originally developed for gravitational lensing of the Cosmic Microwave Background in Cosmology \cite{hu2001mapping, hu2002mass}. The nature of this interaction not only encourages low estimation bias but also enables accurate (and fast) Frequentist mean square error quantification of the nonstationarity characterized by $\bs \theta(\bs x)$. These quadratic estimates are interesting in their own right, as they detect and estimate nonstationarity by probing correlation among Fourier frequencies, the absence of which is the characterizing feature of weak stationarity (by Bochner's Theorem). Moreover, this generalized quadratic estimate can be computed with a Fourier characterization that runs in $\mathcal O(n\log n)$ time and $\mathcal O(n)$ storage when observing the field on a uniform grid of size $n$ in $\Bbb R^d $. Finally, the work presented here partially addresses two unanswered questions that arise in two important bodies of work on nonstationarity: 1) estimating the spatially varying phase in the models analyzed by Dahlhaus \cite{dahlhaus1997fitting,dahlhaus2000likelihood} and 2) how to extend the quadratic estimate of dark matter from the Cosmic Microwave Background to more general nonstationary random fields.

The seminal work of Dahlhaus in the time series literature \cite{dahlhaus1997fitting,dahlhaus2000likelihood}  is  an example of a collection of results that hint at a more general statistical theory of nonstationarity.  Dahlhaus develops asymptotic theory for a particular class of nonstationary random fields modeled by a spatially (or temporally) varying spectral density. These random fields were originally developed for time series \cite{priestley1965evolutionary,priestley1981spectral} and have the form 
\begin{align}
\label{eq: intro nonstat phase model}
Z(\bs x)= \int_{\Bbb R^d} e^{i\bs x\cdot \bs k} A(\bs k, \bs x)\sqrt{C_{\bs k}}\frac{dW_{\bs k}}{(2\pi)^{d/2}}
\end{align} 
where  $C_{\bs k}$ is a spectral density, $dW_{\bs k}$ is an orthogonal increment random complex measure that satisfies $E|dW_{\bs k}|^2 =  d{\bs k}$ (see \cite{Gikhman_Skorokhod_book_v1} for details on random measures) and $A({\bs k},{\bs x})$ represents a spatial (or temporal) modulation of $\sqrt{C_{\bs k}}$. 
Dahlhaus proved consistency results for estimating the squared modulus $|A({\bs k},{\bs x})|^2$ when estimation is done  by maximizing a weighted sum of local likelihood functions. Left unanswered, however, is the question of estimating the phase of $A({\bs k},{\bs x})$. In Section \ref{section: NPhase}, we partly resolve this question by showing that a generalized quadratic estimate can be used to estimate a pure phase modulation $A({\bs k},{\bs x})=\exp({iB(\bs k,\bs x)})$ where $B(\bs k,\bs x)$ is a function taking values in $\Bbb R$ and is a separable function of $\bs x$ and $\bs k$. Indeed, this nonstationary phase model has the local invariance property and, as such, can be accurately estimated (under certain conditions on $B(\bs k,\bs x)$) with the generalized quadratic estimate developed here.

Another important development in the statistical theory of nonstationarity comes from recent gravitational lensing studies of the Cosmic Microwave Background (CMB) \citep{das2011detection,van2012measurement, planck2013lensing, Polarbear2014, planck2015lensing}. In Cosmology, gravitational lensing describes the distortion of photon trajectories due to density fluctuations of intervening dark matter. These density fluctuations affect the CMB observations by introducing small nonstationarities in the original isotropic random field model of the CMB. The state-of-the-art estimator of lensing, the quadratic estimator developed by Hu and Okamoto \cite{hu2001mapping, hu2002mass}, has become an incredibly successful tool for probing the nature of dark matter, understanding cosmic structure and constraining cosmological parameters. What is so surprising about this estimate is that is has small bias. This is due to a delicate cancellation of terms in an infinite Taylor expansion of the lensing effect. Unfortunately there has been no clear explanation as to why this cancellation occurs and whether or not it exists in other models of nonstationarity. In this paper, we argue that this cancellation is due to the fact that the lensing-induced nonstationarity is locally invariant. Moreover, the generalized quadratic estimator developed here extends the lensing estimator to more general settings. Indeed, many of techniques we use to derive the generalized quadratic estimator are inspired by---and closely follow---those used by Hu and Okamoto \cite{hu2001mapping, hu2002mass}.  The point of this paper, in contrast to the work of Hu and Okamoto, is to identify the cause of the delicate Taylor series cancellation and extend the benefits of the lensing estimator to a larger class of nonstationary random fields available to general practitioners of spatial statistics.

@@

% --------------------------------------------------------
%  claim
%---------------------------------------------------------
\begin{claim}[\textbf{Any pair of spectral densities with the same integral are attainable}]
\label{claim any two spec are attainable}
Let $d\geq 1$ be an integer, $t_0 > 0$ be a real number and $C_{\bs k}$, $\tilde C_{\bs k}$ be two spectral densities on $\Bbb R^d$ with finite second moments such that $\sigma^2 = \int_{\Bbb R^d}  C_{\bs k}d\bs k = \int_{\Bbb R^d} \tilde C_{\bs k}d\bs k$. Then there exists a vector field $\bs \eta_{\bs k}\colon \Bbb R^d \rightarrow \Bbb R^d$ which is $L_2(\Bbb R^d)$ integrable with respect to $C_{\bs k}d\bs k$ and  generates a one dimensional curve of spectral measures  $\{\lambda^{(t)}: t\in [0,t_0]\}$, defined by 
\begin{align}
\lambda^{(t)} := (\bs k + t\bs \eta_{\bs k})\,\sharp\, C_{\bs k}
\end{align}
with endpoints $\lambda^{(0)}=C_{\bs k}d\bs k$ and $\lambda^{(t_0)}=\tilde C_{\bs k}d\bs k$,
such that  $\{\lambda^{(t)}:t\in[0,t_0] \}\subset \mathscr C^{C,\bs\eta}$. In particular there exists a nonstationary phase random field model for which  $C_{\bs k}$ and $\tilde C_{\bs k}$ are both locally attainable.
Moreover,  $\{\lambda^{(t)}: t\in [0,t_0]\}$ is a $L_2$-Wasserstein geodesic path within the class of absolutely continuous spectral measures (with total mass $\sigma^2$ and finite second moments) and each measure $\lambda^{(t)}$ has a density $C^{(t)}_{\bs k}$ with respect to Lebesgue measure on $\Bbb R^d$ which (weakly) satisfies 
\begin{align}
\label{div flow for geo}
\partial_t C^{(t)}_{\bs k} +\text{\rm div}(\bs \eta^{(t)}_{\bs k}C^{(t)}_{\bs k}) = 0
\end{align}
for all $t\in [0,t_0]$ where $\bs \eta^{(t)}_{\bs k}:=\bs \eta_{T_t^{-1}(\bs k)}$ and $T_t(\bs k):= \bs k + t\bs \eta_{\bs k}$.
\end{claim}
% --------------------------------------------------------
%  proof
%---------------------------------------------------------
\begin{proof}
By standard optimal transport theory (see \cite{villani2003topics} for example) the assumptions on $C_{\bs k}$ and $\tilde C_{\bs k}$ guarantee the existence of a convex function $\psi_{\bs k}:\Bbb R^d \rightarrow \Bbb R$ such that $\nabla \psi_{\bs k}$ is the optimal transport from $C_{\bs k}d\bs k$ to $\tilde C_{\bs k}d\bs k$.  Let $\bs \eta_{\bs k}:= \frac{1}{t_0}(\nabla \psi_{\bs k}-\bs k)$ so that
\begin{align}
\label{geo eq number 1}
\lambda^{(t)}:=(\bs k + t\bs\eta_{\bs k})\,\sharp\, C_{\bs k} = \big((1-\textstyle\frac{t}{t_0})\bs k + \textstyle\frac{t}{t_0}\nabla\psi_{\bs k}\big)\,\sharp\, C_{\bs k}. 
\end{align}
The particular form of the right hand side of (\ref{geo eq number 1})  implies each measure $\lambda^{(t)}$ has a density $C_{\bs k}^{(t)}$ with respect to Lebesgue measure and the path of measures $\{\lambda^{(t)}\colon t\in[0,t_0] \}$ forms an $L_2$-Wasserstein geodesic with endpoints $C_{\bs k}d\bs k$ and $\tilde C_{\bs k}d\bs k$ at $t=0$ and $t=t_0$ respectively (see Proposition 5.9 in \cite{villani2003topics}). Moreover, $C^{(t)}_{\bs k}$ weakly satisfies (\ref{div flow for geo}) by Theorem 5.34 of \cite{villani2003topics} and the fact that $C^{(t)}_{\bs k} = T_t\,\sharp\, C^{(0)}_{\bs k}$ where $\partial_t T_t(\bs k) = \bs \eta_{\bs k}$.
By  setting $\bs A=t\bs I_d$ in (\ref{eq: local spectral density for nonstationary phase}) on has
\[
\int  \exp\!\big(i(\bs x-\bs y)\!\cdot\! (\bs k + t\bs\eta_{\bs k})\big)\, C_{\bs k}\frac{d\bs k}{(2\pi)^d} 
=\int \exp\!\big(i(\bs x-\bs y)\!\cdot\! \bs\omega\big)\,  C^{(t)}_{\bs \omega}\frac{d\bs \omega}{(2\pi)^d}
\]
which implies that for each $t\in[0,t_0]$ the measure $\lambda^{(t)}$ is a locally attainable spectral measure.
\end{proof}
